% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

% from robert ross thesis
This thesis owes much to the many people who have guided me, supported me, and inspired me throughout the preparation and writing of this work. Below, I attempt to list many of these colleagues, family, and friends, but I cannot hope to thank everyone by name. Thus, upfront, to each and every one I offer my heartfelt thanks.

First and foremost I shall forever be grateful to my academic supervisor, John Bateman, for giving me his guidance, deep insight, patience and a lot of diligent proofreading work when the time finally came. Without him I would never have become a computational linguist and this thesis would not have happened. Similarly, I am very much indebted to my supervisor, Eric Ras, for his kind encouragements, guidance and support right from the beginning starting with PhD proposal writing. Without Eric this thesis could not have started in the first place. 

% from Rebeca's thesis
I believe knowledge is created between people and I would like to thank all those who have shared this process with me. To everyone who has shared a chat over coffee, a talk around the table or a talk at a conference or seminar, thank you. In particular I would like to thank my colleague and friend Muriel Foulonneau, who invited me to join LIST research centre and was always engaging in stimulating discussions. Thanks to Anke Schulz who was the first person I met in need of an SFL parser because she was performing tedious manual corpus annotation. That corpus annotation later became part of Parsimonious Vole evaluation. Thank you to Ela Oren for the work we have done together on corpus annotation, needed in parser evaluation; for inviting me on a short scientific mission to Tel Aviv University; and from whom I have learned about Obsessive Compulsive Disorder. My deep gratitude goes to Daniel Couto Vale who always welcomed me in Bremen and enthusiastically shared his knowledge on Systemic Functional Linguistics. 

There are many friends that have shared this experience with me and I can't thank each of them enough. To any I have inadvertently left out please don't think you are forgotten. A big thank you goes to my friend Mikolaj Podlaszewski with whom I shared lots of thought-provoking philosophical discussions, sometimes fierce debates and who provided me with lots of constructive criticisms. My friend, Andrei Mihalceanu, who unfortunately passed away, has my sincere gratitude for enthusiastic philosophical discussions on language, mind, determinism and entropy. Thanks to Christoph Stahl for his friendship, encouragement and for putting up with me always working on my thesis. 

Even more so than friends, family are there in person and spirit when you need them
most, and that is why they deserve the greatest gratitude of all. A huge thank you to my parents Tamara and Ion Costetchi for unconditional love, encouragement and support. I want to thank my younger brother Cristi. I haven't always been the best big brother
for him, but he has always been there for me. But most of all, I thank Adriana, my beloved wife who gently pushed and encouraged me in the last phase of this thesis and patiently waited for the manuscript to mature. This thesis is for her.  

Finally, I gratefully acknowledge the support of Luxembourg National Research Fund through an AFR PhD grant which made this work possible in the first place. I also want to thank all those who gave me feedback on drafts along the way. However, mistakes, be them of the conceptual or typographic variety, remain mine and mine alone. 

\end{acknowledgements}
